---
title: "Generating Multiple PDF at Once: A use case of Rmarkdown render in a {purrr} loop"
execute: 
  eval: false
---

::: {.callout-note}
This blog post was first published on my previous website in 2019 and has since been lightly revised.
:::

Following on from my earlier post about generating PDFs with R Markdown and a .docx template, a few people asked for more details on how I use R Markdown in day-to-day work. I’ve got a few examples worth sharing, and one I use regularly is batch-generating PDF reports. It’s simple, efficient, and solves a real problem.

Here’s the situation. For some of the modules I teach, I ask students to write short research papers and submit them as PDFs. That’s fine, but annotating PDFs to give feedback is awkward. My workaround is to create a feedback PDF for each student, with comments on their work section by section.

Now, I could manually write and export one PDF per student, but that quickly becomes tedious. Instead, I write my comments in a spreadsheet and then use that to automatically generate all the PDFs at once.

The approach is borrowed from Alison Hill’s presentation ["Made with YAML, strings, and glue. An R Markdown valentine for you](https://alison.netlify.app/rmd-valentine), which is excellent. I’m just applying it to a different case, so if you want a deeper guide, I’d recommend reading her slides. Still, seeing it in a teaching context might be useful.

Let’s get to the code.

## Step 1: Prepare the data

Here are the packages that you need:

```{r}
library(tidyverse)
library(knitr)
```

To keep things light, I’m using a table of TV characters and quotes instead of real student data. Here’s the data frame:

```{r}
ms_1 <- "Sometimes I’ll start a sentence and I don’t even know where it’s going. I just hope I find it along the way."
ms_2 <- "I’m not superstitious, but I am a little stitious."
ms_3 <- "Would I rather be feared or loved? Easy. Both. I want people to be afraid of how much they love me."
lk_1 <- "We have to remember what’s important in life: friends, waffles, and work. Or waffles, friends, work. But work has to come third."
lk_2 <- "What I hear when I’m being yelled at is people caring really loudly at me."
lk_3 <- "There’s nothing we can’t do if we work hard, never sleep, and shirk from all other responsibilities in our lives."
jp_1 <- "Fine, but in protest, I’m walking over there extremely slowly!"
jp_2 <- "I wasn't hurt that badly. The doctor said all my bleeding was internal. That's where the blood's supposed to be."
jp_3 <- "I appealed to their sense of teamwork and camaraderie with a rousing speech that would put Shakespeare to shame."

tribble(
  ~fullname, ~quote_1, ~quote_2, ~quote_3,
  "Michael Scott", ms_1, ms_2, ms_3,
  "Leslie Knope", lk_1, lk_2, lk_3,
  "Jake Peralta", jp_1, jp_2, jp_3
) |>
  kable()
```

Save this as `data_batch.csv`. It’s the input for the batch process.

## Step 2: Build the Rmd template

Create a file called `index.Rmd`. This is the template that will generate one PDF per person. The YAML at the top sets things up:

```{yml}
#| eval: false
---
title: "Batch PDF processing - `r params$fullname` quotes"
output: pdf_document
params:
  fullname: "TV Show Character"
---
```

The key bit is `params:`. It tells RMarkdown to expect a value (in this case, a name) to insert wherever you call `params$fullname`. It also allows you to filter the input data, so each report is personalised.

Within the body of the template, read the CSV and filter it down to just the row that matches `params$fullname`. Then pull in the quotes.

You can do this using inline R code like this:

```{r}
`r filtered_data$quote_1`
```

Or, if you prefer, use [the package {epoxy}](https://github.com/gadenbuie/epoxy) by Garrick Aden-Buie. It’s great for longer texts with lots of variables. In this case, it’s probably overkill, but still a good tool to know.

## Step 3: The batch script

Now, the script that does the heavy lifting. Save this as `_render_batch.R`:

```{r}
data_batch <- read_csv("data_batch.csv")

walk(
  .x = data_batch$fullname,
  ~ render(
    input = "index.Rmd",
    output_file = glue::glue("PDF output - {.x} quotes.pdf"),
    params = list(fullname = {
      .x
    })
  )
)
```

`walk()` takes a list and applies a function to each element. In this case, it runs `render()` for each character name. The params argument tells it which row to pull from the CSV, and `glue::glue()` ensures each output file gets a unique name.

That’s it. No overwriting, no manual labour. One spreadsheet in, a stack of PDFs out.
