---
title: "Time series clustering with Dynamic Time Warping"
author: ''
date: '2018-09-06'
header:
  caption: ''
  image: ''
slug: time-series-clustering-with-dynamic-time-warp
tags: []
categories: []
---

Many solutions for clustering time series are available with R and as usual the web is full of nice tutorials like [Thomas Girke's blog post](http://girke.bioinformatics.ucr.edu/GEN242/pages/mydoc/Rclustering.html), [Rafael Irizarry and Michael Love's book](http://genomicsclass.github.io/book/pages/clustering_and_heatmaps.html), [Andrew B. Collier's blog post](https://datawookie.netlify.com/blog/2017/04/clustering-time-series-data/), [Peter Laurinec's blog post](https://petolau.github.io/TSrepr-clustering-time-series-representations/), [Dylan Glotzer's lecture](http://www.stat.unc.edu/faculty/pipiras/timeseries/Multivariate_6_-_Classification_Clustering_-_Menu.html) or [Ana Rita Marques's module](http://rstudio-pubs-static.s3.amazonaws.com/398402_abe1a0343a4e4e03977de8f3791e96bb.html).
 
Dynamic Time Warping (DTW) is one of these solutions. The main advantage of DTW is the possibility to group time series according their patterns or shapes even if these patterns are not synchronized (lag). 

As far as I know the two main packages which allow time series clustering with DTW are `TSclust` by [Pablo Montero Manso and José Antonio Vilar](https://cran.r-project.org/web/packages/TSclust/index.html) and `dtwclust` by [Alexis Sarda-Espinosa](https://cran.r-project.org/web/packages/dtwclust/index.html). These packages are very simple but powerful tools to analyse time series. However when it comes to analyse real data, I found difficult to understand how the clustering is working. To make this process clearer I'm going to simulate two groups of time series and to check if whether or not the DTW clustering can differentiate them.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## List of packages needed

```{r}
library(dplyr) # data wrangling
library(ggplot2) # grammar of graphics
library(gridExtra) # merge plots
library(ggdendro) # dendrograms
library(gplots) # heatmap
library(tseries) # bootstrap
library(TSclust) # cluster time series
library(dtwclust) # cluster time series with dynamic time warping
```

## Data simulation

Let's imagine two people running a marathon, one had a classic run with a pace increasing with the time and the other had a very bad experience (e.g. "hitting the wall") with a jump in the pace which indicates a significant slow down in the second part of the run. The best is to have real data to analyse but it can be very useful to simulate these pattern in order to assess the clustering efficiency.

A simple way to simulate these time series is to use the `sine` function and to add a random noise in order to make it more credible.

```{r, fig.width=4, fig.height=3, fig.show='hold'}
# classic run
noise <- runif(420) # random noise
x <- seq(1,420) # 42km with a measure every 100m
pace_min <- 5 # min/km (corresponds to fast run)

ts_sim_classic_run <- (sin(x/10)+x/100+noise+pace_min) %>%
  as.ts(.)

ts.plot(ts_sim_classic_run, xlab = "Distance [x100m]", ylab = "Differential pace [min/km]", main = "Example of classic run", ylim=c(0,25))

# wall run
noise <- runif(210) # random noise
x <- seq(1,210) # 21km with a measure every 100m 
pace_min <- 5 # min/km (corresponds to fast run)
pace_wall <- 20 # min/km (corresponds to very slow run) 
ts_sim_part1 <- sin(x/5)+x/50+noise+pace_min
ts_sim_part2 <- sin(x/5)+noise+pace_wall

ts_sim_wall_run <- c(ts_sim_part1,ts_sim_part2) %>%
  as.ts(.)

ts.plot(ts_sim_wall_run, xlab = "Distance [x100m]", ylab = "Differential pace [min/km]", main = "Example of wall run", ylim=c(0,25))
```

A much nicer way would be to use ARIMA with an auto regressive model (AR).

```{r, fig.width=4, fig.height=3, fig.show='hold'}
pace_min <- 5 # min/km (corresponds to fast run)
pace_wall <- 20 # min/km (corresponds to very slow run) 

# classic run
ts_sim_classic_run <- abs(arima.sim(n = 420, mean = 0.001, model = list(order = c(1,0,0), ar = 0.9))) + pace_min

ts.plot(ts_sim_classic_run, xlab = "Distance [x100m]", ylab = "Differential pace [min/km]", main = "Example of classic run", ylim=c(0,25))

# wall run
ts_sim_part1 <- abs(arima.sim(n = 210, model = list(order = c(1,0,0), ar = 0.9))) + pace_min
ts_sim_part2 <- ts(arima.sim(n = 210, model = list(order = c(1,0,0), ar = 0.9)) + pace_wall, start = 211,end =420)

ts_sim_wall_run <- ts.union(ts_sim_part1,ts_sim_part2)
ts_sim_wall_run<- pmin(ts_sim_wall_run[,1], ts_sim_wall_run[,2], na.rm = TRUE)

ts.plot(ts_sim_wall_run, xlab = "Distance [x100m]", ylab = "Differential pace [min/km]", main = "Example of wall run", ylim=c(0,25))
```

## Bootstrap

Now we have two different runs, let's bootstrap them (i.e. replicate with small differences) in order to have two groups of 5 individuals for each run type.

```{r}
ts_sim_boot_classic <- ts_sim_classic_run %>%
  tseries::tsbootstrap(., nb=5, b=200, type = "block") %>%
  as.data.frame(.) %>%
  dplyr::rename_all(funs(c(paste0("classic_",.))))

ts_sim_boot_wall <- ts_sim_wall_run %>%
  tseries::tsbootstrap(., nb=5, b=350, type = "block") %>%
  as.data.frame(.) %>%
  dplyr::rename_all(funs(c(paste0("wall_",.))))

ts_sim_df <- cbind(ts_sim_boot_classic,ts_sim_boot_wall)
```

# Heatmap cluster

Even if I'm a big fan of ggplot2 possibilities, some packages offer efficient ways to compute and plot data. For heatmaps I'm using the `gplots` package which displays time series with dendrograms is a single function. An overlook of all the heatmap possibilities can be found [here](http://www.sthda.com/english/articles/28-hierarchical-clustering-essentials/93-heatmap-static-and-interactive-absolute-guide/)

```{r}
dtw_dist <- function(x){dist(x, method="DTW")}

ts_sim_df %>%
  as.matrix() %>%
  gplots::heatmap.2 (
    # dendrogram control
    distfun = dtw_dist,
    hclustfun = hclust,
    dendrogram = "column",
    Rowv = FALSE,
    labRow = FALSE
  )
```

We can already see an accurate clustering between classic and wall runs but we are interested in DTW analysis so let's implement `TSclust` and `dtwclust` packages.

# DTW cluster

Both `TSclust` and `dtwclust` are following the same steps:

1. Calculating the difference between each time series using the DTW method (but many other distances can be calculated, see for example Montero & Vilar, 2014).
2. Calculating hierarchical cluster analysis over these dissimilarities.
3. Plotting a dendrogram to visually assess the cluster accuracy. The solution to plot the time series with the dendrogram was taken from [Ian Hansel's blog](http://www.hanselsolutions.com/blog/clustering-time-series.html).

## Using `TSclust`

```{r}
# cluster analysis
dist_ts <- TSclust::diss(SERIES = t(ts_sim_df), METHOD = "DTWARP") # note the dataframe must be transposed
hc <- stats::hclust(dist_ts, method="complete") # meathod can be also "average" or diana (for DIvisive ANAlysis Clustering)
# k for cluster which is 2 in our case (classic vs. wall)
hclus <- stats::cutree(hc, k = 2) %>% # hclus <- cluster::pam(dist_ts, k = 2)$clustering has a similar result
  as.data.frame(.) %>%
  dplyr::rename(.,cluster_group = .) %>%
  tibble::rownames_to_column("type_col")

hcdata <- ggdendro::dendro_data(hc)
names_order <- hcdata$labels$label
# Use the folloing to remove labels from dendogram so not doubling up - but good for checking hcdata$labels$label <- ""

p1 <- hcdata %>%
  ggdendro::ggdendrogram(., rotate=TRUE, leaf_labels=FALSE)

p2 <- ts_sim_df %>%
  dplyr::mutate(index = 1:420) %>%
  tidyr::gather(key = type_col,value = value, -index) %>%
  dplyr::full_join(., hclus, by = "type_col") %>% 
  mutate(type_col = factor(type_col, levels = rev(as.character(names_order)))) %>% 
  ggplot(aes(x = index, y = value, colour = cluster_group)) +
  geom_line() +
  facet_wrap(~type_col, ncol = 1, strip.position="left") + 
  guides(color=FALSE) +
  theme_bw() + 
  theme(strip.background = element_blank(), strip.text = element_blank())

gp1<-ggplotGrob(p1)
gp2<-ggplotGrob(p2) 

grid.arrange(gp2, gp1, ncol=2, widths=c(4,2))
```

As expected, The results of `TSclust`show two different groups, one with the classic runs and one with wall runs. However we can see that wall runs are not sorted perfectly according their shape. Let's have a look at `dtwclust` to see if the results are similar.

## Using `dtwclust`

The main asset of `dtwclust` is the possibility to customize the DTW clustering. For more details about all the possibilities, I suggest to have a look at the `dtwclust` package [vignette](https://cran.r-project.org/web/packages/dtwclust/vignettes/dtwclust.pdf).

```{r}
cluster_dtw_h2 <- dtwclust::tsclust(t(ts_sim_df), 
                                    type = "h", 
                                    k = 2,  
                                    distance = "dtw", 
                                    control = hierarchical_control(method = "complete"),
                                    preproc = NULL, 
                                    args = tsclust_args(dist = list(window.size = 5L)))

hclus <- stats::cutree(cluster_dtw_h2, k = 2) %>% # hclus <- cluster::pam(dist_ts, k = 2)$clustering has a similar result
  as.data.frame(.) %>%
  dplyr::rename(.,cluster_group = .) %>%
  tibble::rownames_to_column("type_col")

hcdata <- ggdendro::dendro_data(cluster_dtw_h2)
names_order <- hcdata$labels$label
# Use the folloing to remove labels from dendogram so not doubling up - but good for checking hcdata$labels$label <- ""

p1 <- hcdata %>%
  ggdendro::ggdendrogram(., rotate=TRUE, leaf_labels=FALSE)

p2 <- ts_sim_df %>%
  dplyr::mutate(index = 1:420) %>%
  tidyr::gather(key = type_col,value = value, -index) %>%
  dplyr::full_join(., hclus, by = "type_col") %>% 
  mutate(type_col = factor(type_col, levels = rev(as.character(names_order)))) %>% 
  ggplot(aes(x = index, y = value, colour = cluster_group)) +
  geom_line() +
  facet_wrap(~type_col, ncol = 1, strip.position="left") + 
  guides(color=FALSE) +
  theme_bw() + 
  theme(strip.background = element_blank(), strip.text = element_blank())

gp1<-ggplotGrob(p1)
gp2<-ggplotGrob(p2) 

grid.arrange(gp2, gp1, ncol=2, widths=c(4,2))
```

Now with the cluster are well distributed between classic and wall runs but also inside the clusters where similar shapes appears to be grouped together.

It is possible to modify some argument in order to perform this hierarchical DTW clustering based on z-scores with centroid based on the built-in "shape_extraction" function.

```{r}
cluster_dtw_h2 <- dtwclust::tsclust(t(ts_sim_df), type = "h", k = 2L,
                                    preproc = zscore,
                                    distance = "dtw", centroid = shape_extraction,
                                    control = hierarchical_control(method = "complete"))

hclus <- stats::cutree(cluster_dtw_h2, k = 2) %>% # hclus <- cluster::pam(dist_ts, k = 2)$clustering has a similar result
  as.data.frame(.) %>%
  dplyr::rename(.,cluster_group = .) %>%
  tibble::rownames_to_column("type_col")

hcdata <- ggdendro::dendro_data(cluster_dtw_h2)
names_order <- hcdata$labels$label
# Use the folloing to remove labels from dendogram so not doubling up - but good for checking hcdata$labels$label <- ""

p1 <- hcdata %>%
  ggdendro::ggdendrogram(., rotate=TRUE, leaf_labels=FALSE)

p2 <- ts_sim_df %>%
  dplyr::mutate(index = 1:420) %>%
  tidyr::gather(key = type_col,value = value, -index) %>%
  dplyr::full_join(., hclus, by = "type_col") %>% 
  mutate(type_col = factor(type_col, levels = rev(as.character(names_order)))) %>% 
  ggplot(aes(x = index, y = value, colour = cluster_group)) +
  geom_line() +
  facet_wrap(~type_col, ncol = 1, strip.position="left") + 
  guides(color=FALSE) +
  theme_bw() + 
  theme(strip.background = element_blank(), strip.text = element_blank())

gp1<-ggplotGrob(p1)
gp2<-ggplotGrob(p2) 

grid.arrange(gp2, gp1, ncol=2, widths=c(4,2))
```



Based on `dtwclust` package vignette, it is possible to register a new DTW function adapted to normalized and asymmetric DTW.

```{r}
# Normalized DTW
ndtw <- function(x, y, ...) {
  dtw(x, y, ...,
      step.pattern = asymmetric,
      distance.only = TRUE)$normalizedDistance
}
# Register the distance with proxy
proxy::pr_DB$set_entry(FUN = ndtw, names = c("nDTW"),
                       loop = TRUE, type = "metric", distance = TRUE,
                       description = "Normalized, asymmetric DTW")
# Partitional clustering
cluster_dtw_h2 <- dtwclust::tsclust(t(ts_sim_df), k = 2L,distance = "nDTW")

plot(cluster_dtw_h2)
```

Even if it looks great with `sine` simulated data, it is not very accurate with ARIMA models. Moreover I haven't been able to extract the dendrogram from this last "cluster_dtw_h2" object because of the partitional clustering process but one can be interested in the distance matrix provided in "cluster_dtw_h2" object.

After this short analysis with Dynamic Time Warping, the next steps will be to increase the difference between the time series to check the clustering accuracy and obviously to test it with real data.

